---
title: "Regression in R"
author: "Shashwat Singh"
format: html
editor: visual
---

# Loading Required Packages

```{r}
# Install necessary packages
if (!require(pacman))
  install.packages("pacman")

pacman::p_load(tidymodels,
               tidyverse,
               ranger,
               randomForest,
               glmnet,
               gridExtra)
# Setting up a global theme
theme_set(theme_bw() + theme(legend.position = "top"))

```

## Regression

Regression is a modeling technique for predicting quantitative-valued target attributes. The goals for this tutorial are as follows:

1. To provide examples of using different regression methods for the stock prices dataset.

2. To demonstrate the problem of model overfitting due to correlated attributes in the data.

3. To illustrate how regularization can be used to avoid model overfitting.


# Synthetic Data Generation

```{r}
# This block sets parameters for the analysis.
seed <- 1                   # Set the random seed for reproducibility
numInstances <- 200         # Define the total number of instances in the dataset

# Set seed
set.seed(seed)

# Generating data
X <- matrix(runif(numInstances), ncol=1)
y_true <- -3*X + 1 
y <- y_true + matrix(rnorm(numInstances), ncol=1)

```

- Here we have set seed and created Synthetic data randomly. We are using function which is relating y to x as y=3x+1. Also added gaussian noise.


```{r}
# Plotting the generated data to visualize the linear relationship

ggplot() +
  geom_point(aes(x=X, y=y), color="black") +
  geom_line(aes(x=X, y=y_true), color="blue", linewidth=1) +
  ggtitle('True function: y = -3X + 1') +
  xlab('X') +
  ylab('y')           # Create a scatter plot of the generated data


```


- The points represent the generated data, and the blue line shows the linear relationship between x and y without the additive gaussian noise.



## **Multiple Linear Regression**

Given a training set X,y MLR is designed to learn f(X,w)=X^T^w +w~0~ by minimizing the following loss function given a training set

L(y,f(X,w)) = (i=1 to n)∑ \|\|y~i-~ X~i~w-w~0~\|\|

where w (slope) and w0 (intercept) are the regression coefficients.

Given the input dataset, the following steps are performed:

1\. Split the input data into their respective training and test sets.

2\. Fit multiple linear regression to the training data.

3\. Apply the model to the test data.

4\. Evaluate the performance of the model.

5\. Postprocessing: Visualizing the fitted model.


## 1) Split Input Data into Training and Test Sets

```{r}
# This section splits the data into training and test sets
numTrain <- 20   # number of training instances
numTest <- numInstances - numTrain

set.seed(123) # For reproducibility



# Combine X and y into a tibble for easier data manipulation
data <- tibble(X = X, y = y)

split_obj <- initial_split(data, prop = numTrain/numInstances)



# Split the dataset into training and testing sets
train_data <- training(split_obj)
test_data <- testing(split_obj)

# Extract training and testing data from the split object

# Training sets
X_train <- train_data$X
y_train <- train_data$y

# Test sets
X_test <- test_data$X
y_test <- test_data$y
```

## 2) Fit Regression Model to Training Set

```{r}
# Setting up a linear regression model using the 'lm' engine
lin_reg_spec <- linear_reg() |> 
  set_engine("lm")

# Fit the linear regression model to the training data
lin_reg_fit <- lin_reg_spec |> 
  fit(y ~ X, data = train_data)

# Interpretation: 
# This step involves creating a linear regression model using the training data

```

## 3) Apply Model to the Test Set

```{r}
# This block applies the model to the test data

y_pred_test <- predict(lin_reg_fit, new_data = test_data) |>
  pull(.pred)

# Interpretation: 
# Involves predicting the response variable for the test set using the fitted model

```

## 4) Evaluate Model Performance on Test Set

```{r}
# This block evaluates the performance of the model on the test set

# Create a scatter plot to compare true and predicted values of the response variable for the test set
ggplot() + 
  geom_point(aes(x = as.vector(y_test), y = y_pred_test), color = 'black') +
  ggtitle('Comparing true and predicted values for test set') +
  xlab('True values for y') +
  ylab('Predicted values for y')


# Interpretation: 
# This part assesses how well the model predicts the response variable on new data

```

```{r}
# Organize test data and predictions for model evaluation
eval_data <- tibble(
  truth = as.vector(y_test),
  estimate = y_pred_test
)

# Calculate and display the Root Mean Squared Error (RMSE) and R squared value of the model
rmse_value <- rmse(data = eval_data, truth = truth, estimate = estimate)
r2_value <- rsq(eval_data, truth = truth, estimate = estimate)
```

```{r}
# Output the RMSE and R-squared value to assess the model's goodness of fit

cat("Root mean squared error =", sprintf("%.4f", rmse_value$.estimate), "\n")
cat('R-squared =', sprintf("%.4f", r2_value$.estimate), "\n")
```

## 5) Postprocessing

```{r}
# This block involves postprocessing steps after model evaluation

# Extract and display the slope coefficient from the linear regression model
coef_values <- coef(lin_reg_fit$fit) 
slope <- coef_values["X"]
intercept <- coef_values["(Intercept)"]

```

```{r}
# Output the intercept of the linear regression model
cat("Slope =", slope, "\n")

cat("Intercept =", intercept, "\n")
```

```{r}
# Plot the test data and the predicted linear function, displaying the model's fitted line

ggplot() +
  geom_point(aes(x = as.vector(X_test), y = as.vector(y_test)), color = 'black') +
  geom_line(aes(x = as.vector(X_test), y = y_pred_test), color = 'blue', linewidth = 1) +
  ggtitle(sprintf('Predicted Function: y = %.2fX + %.2f', slope, intercept)) +
  xlab('X') +
  ylab('y')




```
- Created training and test sets. Trained the model on the training data and determined the parameter for the linear regression.